---
title: "Getting Started"
author:
- name: Ahmed Mohamed
  affiliation: Precision & Systems Biomedicine, QIMR Berghofer, Australia
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{lipidr_workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8)
library(knitr)
library(widgetframe)
library(rmarkdown)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.path = "figure/")
knitr::opts_chunk$set(widgetframe_isolate_widgets = TRUE)
knit_print.data.frame = function(x, ...) {
  rmarkdown:::print.paged_df(rmarkdown::paged_table(x))
}
knit_print.DataFrame = function(x, ...) {
  knit_print(as.data.frame(x))
}
knit_print.Iheatmap = function(x, ...) {
  knit_print(frameWidget(iheatmapr::to_widget(x), height = 800))
}
```

# Introduction
`lipidr` implements a series of functions to facilitate inspection,
analysis and visualization of lipidomics datasets. `lipidr`
takes exported Skyline CSV as input, allowing for multiple methods to
be analyzed together.


## Input

`lipidr` represents Skyline files as SummarizedExperiment objects, which can
easily be integrated with a wide variety of Bioconductor packages. 
Sample annotations, such as sample group or other clinical information can be loaded.

## Quality control & visualization
`lipidr` generates various plots, such as PCA score plots and box plots, for quality control of samples and measured lipids. Normalization methods with and
without internal standards are also supported.

## Differential Analysis
Differential analysis can be performed using any of the loaded clinical
variables, which can be readily visualized as volcano plots. A novel lipid
set enrichment analysis (LSEA) is implemented to detect preferential
enrichment of certain lipid classes, chain lengths or saturation patterns.
Plots for the visualization of enrichment results are also implemented.

This vignette provides a step by step guide for downstream analysis of targeted
lipidomics data, exported from Skyline.

# Installation
## From Bioconductor
In R console, type:

```r
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("lipidr")  
```

## From GitHub using devtools
In R console, type:

```r
library(devtools)   
install_github("ahmohamed/lipidr")
```

# Analysis Workflow



# Session information
```{r}
sessionInfo()
```
